
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% LaTex Template for proposals within the              %
% DFG Research Unit Program                            %         
%Planet Formation Witnesses and Probes: Transition Discs
% August 2016                                            %                           
%                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[10pt,fleqn,twoside]{article}

%%%%%%%%%%%%%%%%% GET THE STYLE STUFF %%%%%%%%%%%%%%%%%%%

\input{header.tex}

%%%%%%%%%%%%%%%%% DEFINE THE HEADER TEXT %%%%%%%%%%%%%%%%

\fancyhead[LE,LO]{\slshape
%%%%  Please edit
%
Barbara Ercolano: RU Transition Discs: Coordination Project}
%
%
%%%%%

\begin{document}


\newpage

%%%% PROJECT DESCRIPTION STARTS HERE %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\setcounter{page}{1}

\centerline{\huge\bf\Tcol
%
%
%
%
%%%%  Please edit
%
 Project Z:}
\vspace{1em}

\centerline{\LARGE\bf\Tcol Coordination of the Research Unit}

%
%%%%
%
%
%
%
\vskip1.0cm

%%%%  Please edit

\noindent{\bf Coordinators:}\\
\begin{tabular}{ll}
{\textsf{Main Speaker:}}   & Prof.\ Dr.\ Barbara Ercolano (USM)\\
{\textsf{Vice Speaker:}}   & Prof.\ Dr.\ Cornelis P.~Dullemond (Uni Heidelberg)
\end{tabular}

%%%%  Please edit

\vspace{1em}
\noindent{\bf Requested positions: 50\% secretary position} \\


\section{General}
%
As outlined in the main part of the proposal, in this Research Unit (RU) we
plan several measures to strengthen collaboration and interaction between
the different nodes.  Among these are frequent visits between partner
institutes and the participating nodes, regular Video-conferences, regular
(twice a year) meetings of the members of the RU, and joint invitations of
guests/visitors. Additionally, we plan to organize an international meeting
and invite review speakers from outside, and we plan for a summerschool for
students and interested young scientists within the first year of the RU
funding period.

Within the RU we intend to homogenize the expenditures for travel,
conferences, and guests, to the extent that they are general-purpose for the
RU. Hence, we plan to have a general, central budget from which conference
visits, workshops, short term visits to partner institutes and some visits
to external collaborators are payed. Also guests can be payed out of this
fund. Specialized travel will still be coordinated within each project
separately (as described in each project proposal).

The organization and (financial) management of all these activities requires
a dedicated person dealing with these duties. This person can at the same
time organize the official web-pages for the RU.  For this purpose we apply
for a half-time secretary position at the LMU working with the speaker of
the RU.
%
\section{Funding Request}
%
To estimate our requested funds as listed below we have assumed the
following personnel involved in the RU: 5 PhD students, 4 Postdocs, 7
Principle investigators (PIs), 2 senior collaborators, in total 18 people.

\subsection{Networking}
In the individual project descriptions we have only included travel funds
for direct collaboration between or within (if two nodes are involved in one
project) the individual projects, referring primarily to longer term visits
of approximately a week.

Additionally, we plan here for  occasional short term visits
(collaboration) between scientists, which we plan to organize through this
central budget. Funds for regular meetings of the RU are also organized
centrally. 

For travel within Germany we assume on average 200 Euro per train return ticket
per person and 100 Euro per day for lodging and food. 

We estimate the following amounts (which are totals over 3 years):


\begin{itemize}
\item
\begin{Emphasize}
{\bf Full network meetings.}
\end{Emphasize}\\
We plan for two two-day meetings per year for the whole RU (in total 18
participants; 400~\EUR{}\ per person and meeting). On average we plan to 
invite one international guest per meeting (800~\EUR{}\ travel, 2$\times$100~\EUR{}\ 
lodging and food).\vspace{0.2em}\\
( 400 \EUR{} $\times$ 18 +  1000 \EUR{} ) $\times$ 2 $\times$ 3  \hfill 49,200~\EUR{}\\
\item
\begin{Emphasize}
{\bf Tutorials and Lectures.}
\end{Emphasize}\\
As an extension to the first two full network meetings during the first year, we
intend to add two more days or tutorials and lectures. This means that the
students and postdocs will stay for two more days, as well as two of the PIs
(in addition to the local PI(s)) as lecturers.
\vspace{0.2em}\\
( 200 \EUR{} $\times$ 9 + 200 \EUR{} $\times$ 2 ) $\times$ 2  \hfill 4,400~\EUR{}\\
\item 
\begin{Emphasize}
{\bf Collaboration:} Short term visits to other nodes within the RU.
\end{Emphasize}\\
We plan on average 2 visits for about 2 days per person and year
(i.e.\ in total 400 EUR per person and visit).
\vspace{0.2em}\\
400~\EUR{} $\times$ 2 $\times$ 18 $\times$ 3    \hfill 43,200~\EUR{}\\
\item 
\begin{Emphasize}
{\bf External travel:} Collaboration with non-RU-scientists and visits to conferences.
\end{Emphasize}\\
We plan for the participants of the RU to attend national and
international conferences to present and discuss scientific results.
Successful scientific work requires additional (international) collaborations, and
personal visits between individual scientists.
Therefore, members of the RU will have to visit locations outside the RU. 
For these external activities we estimate on average expenses of (3000~\EUR{})
per project and year.
\vspace{0.2em}\\
3000~\EUR{} $\times$ 8 $\times$ 3    \hfill 72,000~\EUR{}\\
\item 
\begin{Emphasize}
{\bf Guests}: Inviting external (non-RU) scientists for working visits.
\end{Emphasize}\\
We request an average amount for each project per year (in total 3000~\EUR{}) to
be able to invite (international) external collaborators.
\vspace{0.2em}\\
3000~\EUR{} $\times$ 8 $\times$ 3    \hfill 72,000~\EUR{}\\
\item 
\begin{Emphasize}
{\bf International school:} An international summer / winter school on disk physics.
\end{Emphasize}\\
We intend to organize an international summer / winter school on disk physics
during the second year,
with an emphasis on, but not limited to, TDs. We will apply for the Bad Honnef
programme, which would give funding. But to provide additional support for
external international students with little means, we apply for 10,000~\EUR{}. 
\vspace{0.2em}\\
10,000~\EUR{} $\times$ 1              \hfill 10,000~\EUR{}\\
\item 
\begin{Emphasize}
{\bf International conference:} An international conference on TDs.
\end{Emphasize}\\
We intend to wrap up the RU with an international small size conference
on the topic (roughly 60 participants). We apply for 30,000~\EUR{} for this.
\vspace{0.2em}\\
30,000~\EUR{} $\times$ 1              \hfill 30,000~\EUR{}\\
\item 
\begin{Emphasize}
{\bf Consumables.}
\end{Emphasize}\\
For regular expenses of the management (printing costs, postal services, toner,
viewgraphs, copying materials, etc..) we request 1000~\EUR{} per year.
\vspace{0.2em}\\
1000~\EUR{} $\times$ 3              \hfill 3,000~\EUR{}\\
\item 
\begin{Emphasize}
{\bf Publication costs.}
\end{Emphasize}\\
In general we plan to publish our scientific results in journals without
page charge.  However, we would like to stimulate our members to pay 
the additional 700~\EUR{} fee per publication to make the articles 
open access forever. Therefore we ask for 700~\EUR{} per project per
year.
\vspace{0.2em}\\
700~\EUR{} $\times$ 8 $\times$ 3              \hfill 16,800~\EUR{}\\
\item 
\begin{Emphasize}
{\bf Management:} A 50\% part-time secretary position based at the LMU.
\end{Emphasize}\\
For managing the organization and financial matters of the RU, 
organize the web-pages of the RU, as
well as organize the meetings, school and conference, we apply for
a half-time secretary. This position has the following requirements:
Work relatively independently, English language skills and 
basic computer skills. We apply for a E9 position for 50\% part-time.
\vspace{0.2em}\\
53,100~\EUR{} $\times$ 0.5 $\times$ 3
700~\EUR{} $\times$ 8 $\times$ 3              \hfill 79,650~\EUR{}\\
\end{itemize}
 
%%
%\section{Overview of Funding}
%%
%$~$\\[0.3cm]
%The total funds requested for Project Z are summarized in the following table
%(all amounts in \EUR{}):
%\\[0.2cm]
%\centerline{\begin{tabular}{||l|r|r|r||}
%\hline \hline & Year 1 & Year 2 & Year 3 \\ \hline %
%Networking                  & 24.800  & 41,800  & 41,800  \\
%External Travels            & 30.000  & 30.000  & 30.000  \\
%External Guests             & 30.000  & 30.000  & 30.000  \\
%Conference/Summerschool     & 20.000  &   -     & 20.000  \\
%Consumables                 &  1.000  &  1.000  &  1.000  \\
%Publications                &  5.000  &  5.000  &  5.000  \\
%Management                  & 21.000  & 21.000  & 21.000  \\
%\hline
%{\bf Total:} (EUR)          & 131.800  & 128.800 & 148.800 \\
%\hline \hline
%\end{tabular}
%}


\end{document}

